Rheology of Cubic Blue Phases

### 2012-09-06 trunk/papers/bprheo/bprheo.tex
Update bprheo paper
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig@1779 40876441-3496-0410-a613-a2fca35b2ae2

\usepackage{graphicx}
\usepackage{amsmath}

\newcommand{\e}[1]{\times10^{#1}}
\newcommand{\gd}{\dot{\gamma}}

@@

% \affiliation command applies to all authors since the last
% \affiliation command. The \affiliation command should follow the
% other information
% \affiliation can be followed by \email, \homepage, \thanks as well.
\author{O. Henrich}
\email[Corresponding author: ]{o.henrich@ucl.ac.uk}
\affiliation{Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
\affiliation{Centre for Computational Science, University College London, 20 Gordon Street, London WC1H 0AJ, UK}

\author{K. Stratford}
\affiliation{Edinburgh Parallel Computing Centre, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}

\author{P.V. Coveney}
\affiliation{Centre for Computational Science, University College London, 20 Gordon Street, London WC1H 0AJ, UK}

\author{M.E. Cates}
\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}

\author{D. Marenduzzo}
\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh, JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}

@@

% insert suggested PACS numbers in braces on next line
\pacs{47.11.Qr, 47.57.Lj, 47.57.Qk, 61.30.Dk, 61.30.Pq, 61.30.Jf, 83.80.Xz}
% insert suggested keywords - APS authors don't need to do this
%\keywords{}

@@

@@

We start the discussion with BPII. Fig \ref{bp2-med} shows the disclination network 
undergoing an affine transformation under shear. The disclination lines break up and 
reconnect further downstream forming a periodically recurring pattern. 
Apart from the affine transformation the general appearance is very close to that 
of the quiescent equilibrium phase.

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.32\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
of shear-induced transformations in the steady state at 
$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. 
The boundary conditions are applied flow takes place in such as way that network moves 
to the right in the upper half and to the left in the lower half.}
\label{bp2-med}
\end{figure*}

Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
the gradient direction in the direction of vorticity.
Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
making the network travel along the z-direction.
The recurrence period in vorticity direction is exactly six times larger than that of reconnecting in flow direction.

\begin{figure*}[h]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with positive helicity travels rightwards, whereas the one one the right
moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
to reconnect in flow direction.}
\label{bp2-velo}
\end{figure*}

If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation under shear that we
descibed above. Fig. \ref{bp2-high} shows a cut through the director field of BPII at high shear rate.
The entire blue phase network has breaken up into a simple cholesteric phase. The preferred mode of flow is 
a travelling helical wave with the helical axis oriented along the vorticity direction and translational 
invariance in flow and gradient direction. This mode of flow is the preferred one for low Ericksen numbers
before a helix uncoiling transition to a nematic flow aligned state takes place at even higher
shear rates \cite{Rey:1996a, Rey:1996b}.

In our scenario the break-up occurs for shear rates between $\gd=3.12\e{-4}$ and $6.25\e{-4}$.
However, it is reasonable to assume that the critical shear rate depends on thermodynamic and kinetic parameters
such as the effective temperature $\tau$, the chirality $\kappa$ and the rotational diffusion constant $\Gamma$.
Due to a lack of computational resources we have not investigated these aspects any further.

\begin{figure}[h]
\includegraphics[width=0.85\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.85\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom). The images show 1D-cuts as the configuration is translationally invariant in flow and gradient direction. The configuration is a travelling helical wave along the vorticity or z-direction. The colour code indicates the orientation in x-y direction.}
\label{bp2-high}
\end{figure}

@@

\begin{figure*}[h]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
\label{bp2-rheo}
\end{figure*}

@@

The flow behaviour of BPI at low shear rates is is similar deviates significantly from that of BPII in a number of points.

\begin{figure}[h]
\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-600k_run1115.png}\\
\includegraphics[width=0.32\textwidth]{disc-xy-725k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-775k_run1115.png}
\caption{Disclination network of BPI at low shear flow:}
\label{bp1-low}
\end{figure}

First of all it takes considerably longer to reach the steady state with periodically recurring patterns.
This might be because the disclination network of BPI undergoes much more complex flow-induced transformations than that of BPII.
Fig. \ref{fig6} shows a sequence of typical conformations at the same shear rate as Fig. \ref{fig11}, but this time seen along the direction of flow.
A similar affine transformation in flow-gradient plane takes place, which is however obscurred by the transformations shown here. 
The initial and final state in this sequence correspond to half a cycle. 
This can be easily verified by comparing the position of the disclination network at the left or right boundary.
BPI and BPII have exactly the same recurrence period, although it is seemingly four times shorter if the travelling of the network in z-direction is not taken into account.

\begin{figure*}[h]
\includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-367k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-375k_run914.png}\\
\includegraphics[width=0.32\textwidth]{disc-380k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-384k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
\caption{Disclination network of BPI in shear flow: The pictures show a typical sequence of shear-induced transformations in the steady state at $\dot{\gamma}=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$. The horizontal direction is the vorticity direction, whereas the velocity gradient is oriented along the vertical direction.}
\label{bp1-med}
\end{figure*}

The disclination network and the x- and z-component of the flow velocity are depicted in Fig. \ref{fig7}.
Just as in the previous case of BPII inverting the helix in the initial configuration inverts the sense of motion of the network in z-direction.
Equally it leads to an inversion of the velocities, shown here for a particular time steps, but due to the different topology and the complicated conformational changes there is no simple band structure as previously observed for BPII.

\begin{figure*}[h]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run914.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-360k_run922.png}
\caption{Velocity patterns in BPI for positive (left) and negative (right) helicity: The pictures show a characteristic time step of the periodically recurring sequence and velocity vectors $(0,v_y,v_z)$ with the component in flow direction set to zero. The colour code gives the magnitude and sign of the z-component. The vertical and horizontal direction are the gradient and vorticity direction, respectivelty. Flow occurs in such a way that the network moves into (out of) the paper plane in the upper (lower) half of the images.}
\label{bp1-velo}
\end{figure*}

Despite qualitative similarities between the BPI and BPII, the quantitative analysis reveals striking differences.
In Fig. \ref{fig8} we show the effective viscosity $\eta_{eff}$ over time and flow curves $\sigma_{chem}(\dot{\gamma})$. 
BPI exhibits pronounced shear-thinning contrary to BPII, whereas the shape of the curves in no longer symmetric.
In fact on approaching the critical shear rate that led in the case of BPII to the disruption of the network, the curves become increasingly complex but remain fully periodic.
The effective viscosity drops by about an order of magnitude between the lowest and the highest shear rates.
Note that the scale has been adapted to allow for the different magnitudes.
The inset can be directly compared with that of Fig. \ref{fig4}.
Interestingly the difference between minima and maxima of the chemical shear stress is much larger than for BPII and for the lowest shear rate we applied the absolute minimum takes even negative values, which are not shown in Fig. \ref{fig4}.
Our understanding of this phenomenon is an interplay of the elastic forces and the individual order paramter configuration of the liquid crystal.
If the system is close to a local minimum of the free energy but moving away from it the increasing strain pumps elastic energy into the system. 
This results in positive chemical stresses as the system tries to withstand the deformation.
However, if the system has sufficiently transformed and is about to overcome a local maximum, it will be pulled towards the next local minumum, which can lead temporarily to negative chemical stresses. 
The inset shows that the chemical stress decreases monotonically with decreasing flow rate, but does not follow a simple power law dependence.

Nevertheless, the disclination network and order parameter strucutre of both phases looks strikingly different in this high flow rate state.
Fig. \ref{fig10} shows a snapshot along the gradient direction and the direction of vorticity.
Contrary to BPII BPI forms rolls which are advected by the flow and undergo a simple affine transformation similar to that of BPII below the break up point.
They meet in $\pm\pi/2$ disclinations, which are oriented along the z-direction.

\begin{figure}[h]
\includegraphics[width=0.32\textwidth]{dir+z-300k_run916.png}
\includegraphics[width=0.32\textwidth]{dir+z-301k_run916.png}
\includegraphics[width=0.32\textwidth]{dir+z-302k_run916.png}
\caption{Director field of BPI at high shear rate: The pictures show a slice of $L_x\times L_y = 64\times64$ lattice sites along the vorticity direction with the flow / gradient direction oriented along the horizontal / vertical axis. The colour code gives the magnitude of the in-plane component of the director field. The data is translationally invariant along the vorticity direction.}
\label{bp1-high}
\end{figure}

\begin{figure}[h]
\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
\caption{Effective viscosity $\eta_{eff}=\Sigma_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right). The inset shows the flow curve $\sigma_{chem}(\dot{\gamma})$.}
\label{bp1-rheo}
\end{figure}

A Fourier analysis of the time series of the deviatoric chemical stress $\sigma_{eff,xy}$ or alternatively the free energy density $f$ can provide a further insight into the flow behaviour of BPI.
Fig. \ref{fig9} shows the Fourier spectrum of the time series of the chemical stress.
For simplicity we have always doubled the shear rate, so that the first overtone ocoalesces with the fundamental mode of the next highest shear rate. 
At low shear rates we observe a strong contribution of the first harmonic, which is directly related to the recurrence period $T$, shear rate $\dot{\gamma}$ and unit cell size $l_{u}$ via $\omega_0=2\pi/T=2\pi/(l_{u}\dot{\gamma})$.
The magnitude of higher harmonics decreases monotonously with increasing frequency.
Interestingly the inset of Fig. \ref{fig9} shows that at intermediate shear rates the contribution of higher harmonics becomes more pronounced until after a kind of gap and similar break up the fundamental mode dominates again and the former characteristics are restored.
Remarkably the break up in case of BPI occurs in exactly the same flow rate window as BPII.

\begin{figure}[h]
\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
\caption{Spectrum of the time series of the chemical stress over time. The frequency is given in units of $10^2\cdot 2\pi/T$ whereas $T$ is the recurrence period in LB time steps.}
\label{bp1-spec}
\end{figure}

>>>>

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{float}

\newcommand{\e}[1]{\times10^{#1}}
\newcommand{\gd}{\dot{\gamma}}
\newcommand{\com}[1]{{\bf {#1}}}

@@

% \affiliation command applies to all authors since the last
% \affiliation command. The \affiliation command should follow the
% other information
% \affiliation can be followed by \email, \homepage, \thanks as well.

\author{O. Henrich}
\email[Corresponding author: ]{o.henrich@ucl.ac.uk}
\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}
\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}

\author{K. Stratford}
\affiliation{EPCC, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}

\author{P.V. Coveney}
\affiliation{Centre for Computational Science, University College London,\\20 Gordon Street, London WC1H 0AJ, UK}

\author{M.E. Cates}
\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}

\author{D. Marenduzzo}
\affiliation{SUPA, School of Physics and Astronomy, University of Edinburgh,\\JCMB Kings Buildings, Mayfield Road, Edinburgh EH9 3JZ, UK}

@@

% insert suggested PACS numbers in braces on next line
\pacs{47.11.Qr, 47.57.Lj, 47.57.Qk, 61.30.Dk, 61.30.Pq, 61.30.Jf, 83.80.Xz}
% insert suggested keywords - APS authors don't need to do this
\keywords{blue phase, cholesterics, chiral nematics, rheology}

@@

\begin{table*}[htpb]
\begin{tabular}{|c|| c || c |c |c||c| c| c||c| c| c|}
\hline
$\dot{\gamma}$ & $q_0$ & $\bar{v}_{x,min}$ & $\bar{v}_{x,max}$ & $\bar{v}_{x,std}$ & $\bar{v}_{y,min}$ & $\bar{v}_{y,max}$ & $\bar{v}_{y,std}$ & $\bar{v}_{z,min}$ & $\bar{v}_{z,max}$ & $\bar{v}_{z,std}$ \\
\hline
BPI \\
\hline
0.244 & 0.1388 &-15.79 &15.57 &1.04 &-0.89 &0.91 &0.95 &-1.59 &1.19 &1.27 \\
0.488 &0.1388 &-31.11 &31.25 &1.31 &-0.93 &0.98 &1.28 &-1.62 &1.10 &1.40 \\
0.976 &0.1388 &-62.24 &62.37 &2.51 &-1.31 &1.25 &2.68 &-1.24 &0.87 &2.65 \\
1.95 & 0.1388 &-124.65 &124.82 &3.62&  -2.49 &2.50 &4.51 &-1.89 & 1.62 &3.51 \\
3.91 &0.1388 &-248.69 &248.74 &3.90&  -3.31 &3.35 &4.21 &-2.56 & 2.88 &4.39 \\
7.81 &0.1388 &-497.22 &496.16 &4.46 &-6.49 &6.56 &7.99 &-5.31 & 7.46 &6.81 \\ 
15.63 &0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-2.87} & \bf{3.57} &10.54 \\
15.63 &-0.1388 &-992.46 &993.18 &10.36 &-3.08 &3.17 &10.49 &\bf{-3.57} & \bf{2.87} &10.54 \\
31.25 &0.1388 & -1988.36 &1986.71 &16.85 &-4.08 &4.46 &15.94 &-11.37 & 12.16 &19.38\\
62.50 &0.1388 & -4039.41 &3995.3  & 20.57 & -62.63 & 62.13 & 24.68 &-73.52 & 110.76 & 33.26 \\
\hline
BPII \\
\hline
0.244 &0.1963 &-7.75 &7.70 &0.13 &-0.17 &0.17 &0.13 &-0.24 &0.23 &0.19 \\
0.488 &0.1963 &-15.50 &15.40 &0.19 &-0.32 &0.30 &0.21 &-0.43 &0.42 &0.29 \\
0.976 &0.1963 &-31.01 &30.80 &0.33 &-0.60 &0.58 &0.41 &-0.85 &0.79 &0.47 \\
1.95 & 0.1963 &-62.04  &61.60 & 0.97 & -1.09 &1.07 & 0.76 & -1.64 & 1.55 & 0.81\\
3.91 & 0.1963 &-123.93 &123.18 & 1.76 &-1.66 &1.70 & 1.45 &-3.09& 2.73 &1.47\\
7.81 &0.1963  &-247.67 &246.45 & 3.20 &-2.62 &2.71 & 2.67 &-5.78 & 4.77 &2.74\\
15.63 &0.1963 &-494.85 &493.09 & 5.64 &-3.13 &3.31 &4.24 &\bf{-10.00} & \bf{7.66} &4.33\\
15.63 &-0.1963&-494.85 &493.09 & 5.64 & -3.13 &3.31 &4.24 &\bf{-7.66} & \bf{10.00} &4.33\\
31.25 &0.1963 &-988.73 &985.93 &8.96  &-1.79 &1.82 &5.86 &-14.39 & 11.04 &6.35\\
62.50 &0.1963 & -1969.41  & 1969.41 & 3.07 & -1.20 & 1.20 & 0.38 &-0.96 & 1.48 &0.38 \\
\hline
\end{tabular}
\caption{Time-averaged velocity components in shear flow for BPI ($\tau=-0.5, \kappa=1.0$) and BPII ($\tau=-0.5, \kappa=2.0$): Given are minumum and maximum values in $10^{-5}$ lattice units and the maximum standard deviation over the entire time sequence excluding transient dynamics, which means times t$\ge 200000$ for BPI and t$\ge 30000$ in case of BPII.
\label{tab1}}
\end{table*}

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp2.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp2.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\eta\,\gd$ of BPII over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$.}
\label{bp2-rheo}
\end{figure}

We start the discussion with BPII, which shows comparably simpler flow behaviour than BPI.
BPII features simple cubic symmetry and consists of $\pi$-disclination lines that 
intersect and form a characteristic network of nodes.

@@

\subsubsection{Low and intermediate shear rates}

Fig \ref{bp2-med} shows the disclination network 
undergoing an affine transformation under shear. The disclination lines break up and 
reconnect further downstream forming a periodically recurring pattern. 
The recurrence time in flow direction $\tau_F=p/(2\, v_{LE})= 1/\gd$.

\begin{figure}[htpb]
\includegraphics[width=0.4\textwidth]{disc-160k_run902.png}
\includegraphics[width=0.4\textwidth]{disc-164k_run902.png}
\includegraphics[width=0.4\textwidth]{disc-165k_run902.png}
\caption{Disclination network of BPII in shear flow: The pictures show a typical sequence 
of shear-induced transformations in the steady state at 
$\gd=1.56\e{-4}$ and time steps $t=1.60, 1.64,1.65\e{5}$. The velocity gradient 
is oriented along the vertical direction, whereas the horizontal direction is the direction of flow. The boundary conditions impose flow in such a way that network moves 
to the right in the upper half and to the left in the lower half of the picture.}
\label{bp2-med}
\end{figure}

\begin{figure*}[htpb]
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run902.png}
\includegraphics[width=0.495\textwidth]{v_yz-v_z-160k_run903.png}
\caption{Velocity patterns and disclination network in BPII for positive (left) and negative (right) helicity of the 
underlying cholesteric helix: The pictures show velocity vectors $(0,v_y,v_z)$ in x-direction, 
i.e. the velocity has been projected onto a plane perpendicular to the flow direction. 
The vertical and horizontal direction are the gradient and vorticity direction, respectively.
The colour code gives the magnitude and sign of the component in vorticity direction.
The snapshot shows a typical frame during a periodically recurring sequence.
The network on the left with positive helicity travels rightwards, whereas the one one the right
moves leftwards. The recurrence period in vorticity direction is six times longer than the time it takes the network 
to reconnect in flow direction.}
\label{bp2-velo}
\end{figure*}

Apart from the affine transformation the general appearance is very close to that 
of the quiescent equilibrium phase.
Interestingly, while being sheared the disclination network moves perpendicular to the flow and 
the gradient direction in the direction of vorticity.
Contrary to the stick-slip motion that has been observed for sheared blue phases in confined
geometries~\cite{Henrich:2012b} the movement is steady and happens in such a way that the 
positions of breakup and reconnection shown in Fig. \ref{bp2-med} are slightly offset, 
making the network travel along the z-direction.

The recurrence time in vorticity direction $\tau_V=6\tau_F$, 
i.e. it is takes exactly six unit cell displacements in flow directions 
until the network has moved one unit cell in voriticity direction.

\subsubsection{High shear rates}

If the flow rate exceeds a certain critical value BPII ceases to undergo the affine transformation in shear flow that we
descibed above. Fig. \ref{bp2-high} shows a cut through the director field of BPII at high shear rate.
The entire blue phase network has breaken up into a simple cholesteric phase. The preferred mode of flow is 
a travelling helical wave with the helical axis oriented along the vorticity direction and translational 
invariance in flow and gradient direction. This mode of flow is the preferred one for low Ericksen numbers
before a helix uncoiling transition to a nematic flow aligned state takes place at even higher
shear rates \cite{Rey:1996a, Rey:1996b}.

In our scenario the break-up occurs for shear rates between $3.12\e{-4}\le\gd_c\le 6.25\e{-4}$.
However, it is reasonable to assume that the critical shear rate depends on thermodynamic and kinetic parameters
such as the effective temperature $\tau$, the chirality $\kappa$ and the rotational diffusion constant $\Gamma$.
Due to a lack of computational resources we have not investigated these aspects any further.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{dir+y-250k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-253k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-255k_run949.png}
\includegraphics[width=0.495\textwidth]{dir+y-257k_run949.png}
\caption{Director field of BPII at high shear rate: The images show a sequence of time step $t=2.50, 2.53,2.55, 2.57\e{5}$ (from top to bottom) with flow into/out of the page. The configuration is a helical wave that travels along the direction of vorticity. It is translationally invariant in gradient and flow direction.}
\label{bp2-high}
\end{figure}

@@

In BPI the disclination lines are arranged in a body-centered-cubic array.
Contrary to BPII, they are well separated and do not intersect. 
There is strong evidence that these topological differences result in 
a very different rheological response of BPI. In fact BPI exhibits some 
peculiarities that are absent in BPII.

Just as before in a kind of overview we present rheological key quantities first and 
will address the detail aspects further below. 
Fig. \ref{bp1-rheo} shows the apparent viscosity over time (left graph) and 
strain (right graph) for intermediate and high shear rates. In case of BPI the oscillatory patterns
are no longer symmetric as it takes more time to build up stress and reach a maximum than to 
relax. Shear-thinning is much more pronounced than for BPII. At the lowest shear rate $\gd=9.76\e{-6}$
the apparent viscosity varies by almost a factor 8 during one cycle.
The decrease in apparent viscosity is monontonous for decreasing shear rate, but does not 
follow a simple power law.

The inset of shows the flow curve, i.e. shear stress over shear rate.
The data points are averages over one cycle, whereas the error bars indicate the maximum and
minmum values during one cycle. Depending on the shear rate three different regimes of flow emerge:
At high shear rates the network breaks up similar to BPII. The director field in this state
is effectively two-dimensional and translationally invariant along the neutral direction.
At intermediate shear rates the BPI network flows with periodically recurring conformations.
These complex flow-induced conformations are accompanied by large fluctuations that increase 
in intensity for decreasing shear rate. At low shear rates the minimum shear stress contribution
of the liquid crystal during one cycle is actually negative. 
Initially, the network undergoes an affine transformation, but it becomes unstable at larger stages. 
Finally it dissolves into an amorphous configuration that has separated disclination lines. This 
end state is the same for all shear rates below a critical threshold, which was below $\gd=9.76\e{-5}$.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp1.pdf}
\includegraphics[width=0.495\textwidth]{stress_vs_strain_bp1.pdf}
\caption{Apparent viscosity $\eta_{app}=\Pi_{xy}/\dot{\gamma}$ of BPI over time (left) and strain (right). The inset shows the flow curve $\bar{\Pi}_{xy}(\gd)$. Three different regimes can be identified: amorphous BP network with yield stress (triangles), flow with recurring oscillatory patterns (open circles), breakup (solid circle). The error bars give the minimum and maximum shear stress during one cycle.}
\label{bp1-rheo}
\end{figure}

The Fourier analysis of the time-dependent deviatoric stress $\Pi_{xy}$ 
offers another perspective on this intricate flow behaviour at high and intermediate shear rates. 
The spectrum $X_\omega$ according to Eq. \ref{spectrum} is shown in Fig. \ref{bp1-spectrum}.
We observe a strong contribution of the first harmonic, which is directly related to the 
recurrence period $T$, shear rate $\dot{\gamma}$ and unit cell size $l_{u}$ 
via $\omega_0=2\pi/T=2\pi/(l_{u}\dot{\gamma})$.
As the shear rates we applied always differ by a factor 2 the Fourier coefficients of the 
individual runs with different shear rate coincide in such a way that the
the fundamental mode is the same frequency as the first overtone of the next lower shear rate
or the third overtone of the next-to-next lower shear rate 
(viz. peak positions at $\gd=7.81, 3.91$ and $1.95\e{-5}$).
Moreover, the magnitude of higher Fourier modes decreases monotonously.   
However, the inset shows that this characteristic frequency relation is violated 
for shear rates above $\gd=7.81\e{-5}$.
Although the velocity of the primary flow component in x-direcition, listed in Tab. \ref{tab1},
 is very well in accordance with the imposed shear rate, the fundamental mode appears 
now at the same frequency.
The contribution of higher Fourier modes increases and even exceeds that of
the fundamental mode, which gives the shear stress generally a more jittering appearance.
At the highest imposed shear rate where the breakup occurs the Fourier spectrum 
becomes simple again and features a dominating fundamental mode at the 
expected frequency just as described above for $\gd\le7.81\e{-5}$.
The breakup occurs at exactly the same flow rate as those of BPII, 
suggesting that this can be traced back to the rotational dynamics 
of the order parameter.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI at different shear rates: The main picture shows the frequency shift of the Fourier components in agreement with the imposed shear rate. The inset depicts the same analysis at higher shear rate, where the frequency shift is no longer observed. Instead the spectral contribution of higher modes grows until finally the breakup takes place. The frequency is given in units of $10^2\cdot 2\pi/T$ with $T$ as period of one cycle in LB time steps.}
\label{bp1-spectrum}
\end{figure}

\subsubsection{Low shear rates}

The rheological response of BPI at low shear rate which exhibits a
dynamic yield stress is strikingly different from that of BPII.
An explanation for this strange behaviour can be found by 
looking more closely at the average deviatoric stress and 
free energy density, as shown in Fig. \ref{bp1-fe-yield}.
When the quiescent and equilibrated BPI network begins to flow
the shear stress increases steeply and goes through a global maximum.
Shortly after it goes negative (the total shear stress is still
positive as the backgound viscosity has been subracted) and positive
again, forming a complete cycle. 
This cycle is repeated for a couple of times before 
the amorphous configuration emerges and any regular pattern
disappears.
The negative branch in the deviatoric stress is connected
to a local maximum followed by a local minimum in the average free energy.
Hence, the region of negative stress can be 
regarded as a phase during which elastic forces excert a 
'pull' onto the network while it tries to reach a more 
favourable configuration. However, the low flow rate
causes this unstable intermediate state to be exposed to these
pulling elastic forces for a relatively long time, long enough to 
create small distortions which finally break the symmetry of the
flow-induced configuration.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_fe_yield_bp1.pdf}
\caption{Average deviatoric stress and free energy density of BPI at low shear rate: The negative branch in the stress is related to a local maximum and a following local minimum in free energy density. This creates unstable conditions that lead to a dissolving regular network and an amorphous state which features a yield stress.}
\label{bp1-fe-yield}
\end{figure}

We want to explain these differences on the basis of a typical 
appearance of BPI at low flow rate, which is depicted in Fig. \ref{bp1-low}.
Three different stages can be distinguished. 
With the onset of shearing the disclination lines 
in BPI get more and more squeezed together which is incommensurable with 
the defect topology of this blue phase. 
Consequently, the network adopts a flow-induced conformation which consists 
of intertwined helices that stretch during the affine transformation.
This helical conformation that emerges already at strains $\gamma\simeq1$ 
is shown in Fig. \ref{bp1-low} (top row).
At a particular point in the cycle a non-helical configuration is
observed (bottom left) before the network quickly returns
to its helical state.
There is no perceivable movement of the network in vorticity 
direction at any stage of the cycle.

\begin{figure*}[htpb]
\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
\caption{Snapshots of BPI disclination network at low shear rate ($\dot{\gamma}=4.88\e{-6}$): Depicted is the transition from flow-induced, intertwined helices that undergo a recurring affine transformation (top left and centre at time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) to a distorted state (bottom centre at time step $t=1.7\e{6}$) and finally to an amorphous network (bottom right at $t=2.2\e{6}$). All pictures have the vie along the vorticity direction apart from the one at the top centre which shows the network in gradient direction.}
\label{bp1-low}
\end{figure*}

However, this mode of flow proves unstable as after a few cycles 
distortions appear which lead to further destabilisation.
Finally, the network loses any residual order and 
transforms into an amorphous state with virtually constant
yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LU.

\subsubsection{Intermediate shear rates}

Intermediate shear rates in the sense of this study of BPI are characterised 
by the missing negative branch in the deviatoric stress. It is unclear 
why this feature is missing at higher flow rates. A possible explanation is that 
above a certain flow rate the network simply skips the comparably small 
window of negative stress in Fig. \ref{bp1-fe-yield} and finds a quicker 
route to get back on the positive branch.

Fig. \ref{bp1-med} shows snapshots of BPI in steady shear flow. 
Contrary to BPII the flow-induced conformations of BPI differ significanly 
from the appearance of BPI at equilibrium. The depicted 
state is rather surprisingly still kinematically connected to the equilibrium state.
When the shear flow is switched off the flow-induced configuration reverts
to its original configuration of a BPI network at rest.
%%%%%%%%%%%%
%The recurrence time in flow direction is $\tau_F=L_x/(2\, v_{LE})=L_x/(2\,\gd\, \Delta y_{LE})= 4/\gd$.
%%%%%%%%%%%
After each cycle the position of the network is displaced with respect to the beginning of the cycle.
However, neither the magnitude nor the direction of the offset obey a clear rule,
are different and non-integer for every shear rate in this regime of flow. 
This suggests the described phenomenon constitutes an instance of rheochaos in this 
regime of liquid crystalline flow.

\begin{figure*}[htpb]
\includegraphics[width=0.32\textwidth]{disc-365k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-367k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-375k_run914.png}\\
\includegraphics[width=0.32\textwidth]{disc-380k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-384k_run914.png}
\includegraphics[width=0.32\textwidth]{disc-389k_run914.png}
\caption{Disclination network of BPI at intermediate flow rates: The pictures show a typical sequence of shear-induced transformations in the steady state at $\gd=1.56\e{-4}$ and time steps $t=3.65, 3.67,3.75,3.80,3.84,3.89\e{5}$ with the flow into/out of the page. The network is displaced during every cycle, but the magnitude and direction of the offset appear to be chaotic.}
\label{bp1-med}
\end{figure*}

Fig. \ref{bp1-velo} depicts the secondary velocity components $v_y$ and $v_z$ in the same fashion as Fig. \ref{bp2-velo} for BPII. Again, the orientation of the components are as mirror images of each other if the helicity of the cholesteric changes from left- to right-handed.
The emerging pattern is different from the one observed before in BPII, but agrees with
the previous result in the sense that secondary velocities are largest in the regions
where large conformational changes take place.
